Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-12 paper/paper.tex

\newcommand{\SPF}{\texttt{SPF}}
\newcommand{\Var}{\texttt{Var}}
\newcommand{\K}{\texttt{K}}
\newcommand{\map}{\texttt{map}}
\newcommand{\roll}{\texttt{roll}}
\newcommand{\fold}{\texttt{fold}}
\newcommand{\inl}{\texttt{inl}}
\newcommand{\inr}{\texttt{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\stringg}{\texttt{String}}
\newcommand{\charg}{\mathtt{Char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

For reasons of brevity, we present only a selection of cases in
\Cref{fig:semantics}. The full details are in the appendix.\max{TODO:
  put the full details in the appendix}.
\max{TODO: how to describe $\mu$ succinctly?}\max{make sure $\lto$ and $\tol$ are right}
\begin{figure}
  \begin{align*}
    \sem{c}\gamma\,w &= \{ c | w = c\}\\
    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
    \sem{\SPF\,X}\gamma &= \{ F : \textrm{Functor}(\textrm{Grammar}^{\sem{X}\gamma}, \textrm{Grammar}) \pipe F \textrm{ preserves $\omega$-colimits.} \}\\
    \sem{\el(F)}\gamma\,G &= \sem{A}\gamma\,G\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:adamek}}
  \end{align*}
  \caption{Grammar Semantics}
\end{figure}

\max{TODO: prose}

\max{TODO: spelling}
\begin{theorem}[Adamek]
  If a functor $F : \mathcal C \to \mathcal C$ preserves
  $\omega$-colimits and $\mathcal C$ has an initial object and
  $\omega$-colimits, then $F$ has an initial algebra given by the
  colimit of the chain
  \[ 0 \to F(0) \to F(F(0)) \to \cdots \]
\end{theorem}

@@

>>>>

\newcommand{\Subst}{\textrm{Subst}}
\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\String}{\textbf{String}}
\newcommand{\stringg}{\texttt{String}}
\newcommand{\charg}{\mathtt{Char}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

For reasons of brevity, we present only the cases related to the
linear types in \Cref{fig:semantics}. The full details are in the
appendix.\max{TODO: put the full details in the appendix}.  The
grammar for a literal $c$ has a single parse precisely when the input
string consists of the single character. The grammar for the unit
similarly has a single parse for the empty string.  A parse of the
tensor product $A \otimes B$ consists of a \emph{splitting} of the
empty string into a prefix $w_1$ and suffix $w_2$ along with an $A$
parse of $w_1$ and $B$ parse of $w_2$. A parse of $\bigoplus_{x:X} A$
is a pair of an element of the set $X$ and a parse of $A(x)$, while
dually a parse of $\bigwith_{x:X} A$ is a \emph{function} taking any
$x:X$ to a parse of $A(x)$. A $w$-parse of $A \lto B$ is a function
that takes an $A$ parse of some other string $w'$ to a $B$ parse of
$ww'$, and $B \tol A$ is the same except the $B$ parse is for the
reversed concatenation $w'w$.
%
The set $\uparrow A$ is the set of parse for the empty string for
$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
equivalent to the set of parse transformers:
\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
Next, a parse in the equalizer $\equalizer{a : A}{f}{g}$ is defined as
a parse in $\sem{A}$ that is mapped to the same parse by $f$ and
$g$. Recall that $\sem{f}\gamma$ and $\sem{g}$ are elements of
$\sem{\uparrow(A \lto B)}$, i.e., parse transformers.

The universe $L$ of linear types is interpreted as the set of all
(small) grammars.

\begin{figure}
  \begin{align*}
    \sem{c}\gamma\,w &= \{ c | w = c\}\\
    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
    \sem{\equalizer{a:A}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
    \sem{L}\gamma &= \Grammar_0\\
    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma) \tag{\Cref{thm:indexed-container}}\\
  \end{align*}
  \caption{Grammar Semantics}
\end{figure}

The most complex part of the semantics is the interpretation of
strictly positive functors and indexed inductive linear types.  We
interpret a strictly positive functor as a \emph{dependent polynomial
functor} on the category of sets, also sometimes called an
\emph{indexed container} \cite{dep-poly,indexed-container}.
\begin{definition}
  Let $I$ and $O$ be sets. A (dependent) \emph{polynomial} (of sets)
  from $I$ to $O$ consists of a set of shapes $S$, a set of positions
  $P$ and functions $I \leftarrow P \to S \to O$. The \emph{extension}
  of a polynomial is a functor $\Set/I \to \Set/O$ defined by
  \max{TODO: fill in details}.

  A dependent polynomial functor from $I$ to $O$ is a functor $\Set/I
  \to \Set/O$ that is nautrally isomorphic to the extension of a
  dependent polynomial from $I$ to $O$.
\end{definition}
With this interpretation, $\el(F)$ and $\map(F)$ are just interpreted
as the action of the functor on objects and morphisms,
respectively. We interpret the constructors $\mathsf{K},\Var$, etc. on
functors in the obvious way that matches the definitional behavior of
$\el$ and $\map$. The non-trivial part of the construction is
verifying that such functors are constructing corresponding dependent
polynomials. The details are tedious but straightforward extension of
prior work on dependent polynomials and indexed containers and we have
verified the construction in Agda.

The reason we use dependent polynomials functors on sets is that it is
well-known that their initial algebras exist
\cite{initial-algebras-of-dep-poly}. Additionally, initial algebras
for dependent polynomials are straightforwardly constructed in Agda as
an inductive type of ``IW'' trees, which is available as a
construction in the cubical library of Agda \cite{iw,cubical-library}.
\begin{theorem}
  Every dependent polynomial functor from $I$ to $I$ has an initial
  algebra.
\end{theorem}
Then an element $F \in \sem{X \to \SPF\,X}\gamma$ is an
$\sem{X}\gamma$-indexed family of polynomial functors from
$\sem{X}\gamma\times \Sigma^*$ to $\Sigma^*$, and taking the product
of these constructs a polynomial functor from $\sem{X}\gamma\times
\Sigma^*$ to itself. Then $\sem{\mu\,F}\gamma$ is defined to be the
initial algebra of this functor, and the initial algebra structure is
used to interpret $\roll, \fold$ and the corresponding axioms.

@@

\begin{definition}
  The set of (non-linear) substitutions $\gamma \in
  \textrm{Subst}(\Gamma,\Gamma')$ where $\Gamma \isCtx$ and $\Gamma'
  \isCtx$ is defined by recursion on $\Gamma$:
  \begin{align*}
    \textrm{Subst}(\Gamma,\cdot) &= \{\cdot \}\\
    \textrm{Subst}(\Gamma,\Gamma',x:A) &= \{ (\gamma,M/x) \pipe \gamma \in \textrm{Subst}(\Gamma,\Gamma') \wedge \Gamma \vdash M : A[\gamma] \}
  \end{align*}
  simultaneously with an action of substitution on types, terms,
  etc. in the standard way.

  \max{do we need identity and composition of substitutions and their functoriality?}
  %% By induction we can define an identity substitution
  %% $\textrm{Subst}(\Gamma,\Gamma)$ which maps each variable to
  %% itself. We write this substitution simply as $\Gamma$.
  %% \begin{align*}
  %%   \cdot \circ \gamma' &= \cdot\\
  %%   (\gamma,M/x) \circ \gamma' &= (\gamma\circ \gamma') , M[\gamma']/x
  %% \end{align*}

  It is straightforward, but laborious to establish that all forms in
  the type theory that are parameterized by a non-linear context
  $\Gamma$ support an admissible action of a substitution $\gamma \in
  \Subst(\Gamma',\Gamma)$:
  \begin{mathpar}
    \inferrule
    {\Gamma \vdash X \isTy}
    {\Gamma' \vdash X[\gamma] \isTy}

    \inferrule
    {\Gamma \vdash X \isSmall}
    {\Gamma' \vdash X[\gamma] \isSmall}

    \inferrule
    {\Gamma \vdash X \equiv Y}
    {\Gamma' \vdash X[\gamma] \equiv Y[\gamma]}

    \inferrule
    {\Gamma \vdash M : X}
    {\Gamma' \vdash M[\gamma] : X[\gamma]}

    \inferrule
    {\Gamma \vdash M \equiv N : X}
    {\Gamma' \vdash M[\gamma] \equiv N[\gamma] : X[\gamma]}

    \inferrule
    {\Gamma \vdash \Delta \isLinCtx}
    {\Gamma' \vdash \Delta[\gamma] \isLinCtx}

    \inferrule
    {\Gamma \vdash A \isLinTy}
    {\Gamma' \vdash A[\gamma] \isLinTy}

    \inferrule
    {\Gamma \vdash A \equiv B}
    {\Gamma' \vdash A[\gamma] \equiv B[\gamma]}

    \inferrule
    {\Gamma;\Delta \vdash e : A}
    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] : A[\gamma]}j

    \inferrule
    {\Gamma;\Delta \vdash e \equiv f : A}
    {\Gamma'; \Delta[\gamma] \vdash e[\gamma] \equiv f[\gamma] : A[\gamma]}j
  \end{mathpar}

  %% Further, these all satisfy identity and composition equations M[\Gamma] = M and M[\gamma][\gamma'] = M[(\gamma \circ \gamma')]
\end{definition}

\begin{definition}
  Let $\Gamma \vdash \Delta \isLinCtx$ and $\Gamma \vdash \Delta'
  \isLinCtx$. The set of linear substitutions $\Subst(\Delta',\Delta)$
  is defined by recursion on $\Delta$:
  \begin{align*}
    \Subst(\Delta',\cdot) &= \{ \cdot \pipe \Delta' = \cdot \}\\
    \Subst(\Delta',(\Delta,a:A)) &= \{ (\delta, e/a) \pipe \delta \in \Subst(\Delta_1, \Delta) \and \Delta_2 \vdash e : A \and \Delta' = \Delta_1,\Delta_2\}
  \end{align*}

  Given substitutions $\delta_1 \in \Subst(\Delta_1', \Delta_1)$ and
  $\delta_2 \in \Subst(\Delta_2', \Delta_2)$, we can define a
  substitution $\delta_1,\delta_2 \in
  \Subst((\Delta_1',\Delta_2'),(\Delta_1,\Delta_2))$. Furthermore, for
  any substitution $\delta \in \Subst(\Delta,(\Delta_1,\Delta_2))$, we
  can deconstruct $\delta = \delta_1,\delta_2$ with $\delta_1 \in
  \Subst(\Delta_1', \Delta_1)$ and $\delta_2 \in \Subst(\Delta_2',
  \Delta_2)$.
  %% Again for any $\Gamma \vdash \Delta$, we can define an identity
  %% substitution mapping each variable to itself. Additionally we can define composition of 
\end{definition}

\begin{definition}
  Given any $\Gamma ; \Delta \vdash e : A$ and $\delta \in
  \Subst(\Delta',\Delta)$, we define the action of the substitution on
  $e$ as follows, frequently using the inversion principle to split
  the substitution into constituent components.
  \begin{align*}
    a[e/a] &= e\\
    (e_1,e_2)[\delta_1,\delta_2] &= (e_1[\delta_1], e_2[\delta_2])\\
    (\letin {(a , b)} e {e'})[\delta_1,\delta_2,\delta_3] &= \letin{(a,b)} {e[\delta_2]}{e'[\delta_1,a/a,b/b,\delta_2]}\\
    ()[\cdot] &= ()\\
    \letin{()} e {e'}[\delta_1,\delta_2,\delta_3] &= \letin{()} {e[\delta_2]}{e'[\delta_1,a/a,b/b,\delta_2]}\\
    (\lamblto {a} e)[\delta] &= \lamblto a {e[\delta,a/a]}\\
    (\applto {e'} {e})[\delta_1,\delta_2] &= \applto {e'[\delta_1]} {e[\delta_2]}\\
    (\lamblto {a} e)[\delta] &= \lamblto a {e[\delta,a/a]}\\
    (\applto {e'} {e})[\delta_1,\delta_2] &= \applto {e'[\delta_1]} {e[\delta_2]}\\
    (\lambtol {a} e)[\delta] &= \lamblto a {e[a/a,\delta]}\\
    (\apptol {e'} {e})[\delta_1,\delta_2] &= \apptol {e'[\delta_1]} {e[\delta_2]}\\
    (\dlamb x e)[\delta] &= \dlamb x {e[\delta]}\\
    (e\,.\pi\,M)[\delta] &= (e[\delta]\,.\pi\,M)\\
    (\sigma\,M\,e)[\delta] &= \sigma\,M\,e[\delta]\\
    (\letin{\sigma\,x\,a}{e}{e'})[\delta_1,\delta_2,\delta_3] &= \letin{\sigma\,x\,a} {e[\delta_2]}{e'[\delta_1,\delta_3]}\\
    (\equalizerin{e})[\delta] &= \equalizerin{e[\delta]}\\
    (\equalizerpi{e})[\delta] &= \equalizerpi{e[\delta]}
  \end{align*}
  By induction on linear term and equality judgments, we establish the
  following admissible rules for $\delta \in \Subst(\Delta',\Delta)$:
  \begin{mathpar}
    \inferrule
    {\Gamma; \Delta \vdash e : A}
    {\Gamma; \Delta' \vdash e[\delta] : A}

    \inferrule
    {\Gamma; \Delta \vdash e \equiv f : A}
    {\Gamma; \Delta' \vdash e[\delta] \equiv f[\delta'] : A}
  \end{mathpar}

  \max{identity and composition?}
\end{definition}

\section{Denotational Semantics}

\begin{definition}
  We define the unit grammar
  \[ I\,w = \{ ()\pipe w = \varepsilon \}\]
  and given grammars $A$ and $B$ we define their tensor product
  \[ (A \otimes B)\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in A\,w_1 \wedge b \in B\,w_2 \} \]
  This extends to a monoidal structure on the category of grammars. Note that we defined in the semantics of linear types $\sem{I}\gamma = I$ and $\sem{A \otimes B}\gamma = \sem{A}\gamma \otimes \sem{B} \gamma$
\end{definition}

\begin{definition}[Denotation of Linear Contexts]
  The semantics of linear contexts $\Gamma \vdash \Delta \isLinCtx$ is
  defined as follows:
  \max{make sure this aligns with our interpretation of $\lto$/$\tol$}
  \begin{align*}
    \sem{\cdot}\,\gamma &= I \\
    \sem{\Delta,x:A}\,\gamma &= \sem{\Delta}\,\gamma \otimes \sem{A}\gamma \\
  \end{align*}
\end{definition}

We define denotations of linear terms in
\Cref{fig:linear-term-semantics}. Note that the denotations interpret
typing derivations, not raw terms, as the data of how contexts are
split is needed in order to construct the correct associator
functions.
\max{todo: describe associators, define denotations, prove axioms}
\begin{figure}
  \begin{mathpar}
  \inferrule
  {\Delta_1 \vdash e_1 : A_1 \and \Delta_2 \vdash e_2 : A_2}
  {\sem{(e_1,e_2)}\gamma\,w\,\delta
    = \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}
  }

  \inferrule
  {\Delta_2 \vdash e : A \otimes B \and \Delta_1,a:A,b:B,\Delta_3 \vdash e' : C}
  {\sem{\letin{(a,b)}{e}{e'}}\gamma\,w\,\delta =
    \letin{(w_1,w_2w_3,\delta_1,(w_2,w_3,\delta_2,\delta_3))}{m_{\Delta_1,\Delta_2,\Delta_3}(\delta)}
    \letin{(w_{2,1},w_{2,2},a,b)}{\sem{e}\gamma\,w_2\,\delta_2}
    {\sem{e'}\gamma\,w\,(w_1,w_{2,1}w_{2,2}w_3, \delta_1, (w_{2,1},w_{2,2}w_3, a, (w_{2,2},w_3, b, \delta_3)))}
    }
\end{mathpar}
  \caption{Denotations of linear terms}
\end{figure}

\begin{definition}
  For any $\Gamma \vdash \Delta_1,\Delta_2 \isLinCtx$ and $\gamma \in
  \sem{\Gamma}$ there is an isomorphism $m_{\Delta_1,\Delta_2} :
  \sem{\Delta_1, \Delta_2}\gamma \cong \sem{\Delta_1}\gamma \otimes
  \sem{\Delta_2}\gamma$.

  This can be extended to a sequence of contexts of any length.
\end{definition}

\begin{lemma}[Equations for $\otimes$]

\end{lemma}

\begin{lemma}[Equations for $I$]

\end{lemma}

\begin{lemma}[Equations for $\lto$]

\end{lemma}

\begin{lemma}[Equations for $\tol$]

\end{lemma}

\begin{align*}
  \sem{(e_1,e_2)}\gamma\,w\,\delta &=
  \letin{(w_1,w_2,\delta_1,\delta_2)} {m(\Delta_1,\Delta_2)(\delta)} {(w_1,w_2,\sem{e_1}\gamma\,w_1\,\delta_1, \sem{e_2}\gamma\,w_2\,\delta_2)}\\
\end{align*}
